---
title: "Response to Reviewers — drug-response prediction (Fig 8) analyses"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 4.8)
suppressPackageStartupMessages({ library(tidyverse); library(readxl); library(patchwork); library(ggpubr) })
```

<style>
.rev   { background:#fff6d5; border-left:5px solid #e0b400; padding:10px 14px; margin:14px 0; }
.resp  { background:#eef6ff; border-left:5px solid #4575b4; padding:10px 14px; margin:14px 0; }
.todo  { background:#fdecea; border-left:5px solid #c0392b; padding:10px 14px; margin:14px 0; }
.note  { color:#666; font-size:90%; }
.tag   { display:inline-block; background:#2e8b57; color:#fff; font-size:78%; font-weight:600;
         padding:1px 7px; border-radius:3px; letter-spacing:.3px; }
</style>

```{r loaders}
outdir <- "/lustre/home/huynhk4/BioinformaticsCore/Project 2/results/results_v2/Fig8_rerun_artifact"
ddir   <- "/lustre/home/huynhk4/BioinformaticsCore/Project 2/data_version_2"
drop_pdx <- "VCU-MC-101"

sample_rename <- c("VCUBC003_052223"="VCU-BC-003")
clean_samples <- function(x){x<-gsub("\\.","-",x);x<-sub("_.*","",x);ix<-x%in%names(sample_rename);x[ix]<-sample_rename[x[ix]];x}
load_obs <- function(p){d<-as.data.frame(read_excel(p));colnames(d)[2]<-"Target"
  w<-d|>dplyr::select(-Target)|>tidyr::pivot_longer(-Drug,names_to="Sample",values_to="V")|>
    tidyr::pivot_wider(names_from=Drug,values_from=V)|>as.data.frame();rownames(w)<-clean_samples(w$Sample);w$Sample<-NULL;as.matrix(w)}
rp <- function(f){p<-file.path(outdir,f); if(!file.exists(p)) return(NULL)
  x<-as.data.frame(readr::read_csv(p,show_col_types=FALSE));rownames(x)<-clean_samples(x[[1]]);x[[1]]<-NULL;as.matrix(x)}
rd <- function(f){p<-file.path(outdir,f); if(!file.exists(p)) return(NULL); readr::read_csv(p,show_col_types=FALSE)}

O10 <- load_obs(file.path(ddir,"20251113_All_U54_10uM_inhibition.xlsx"))
O10=O10[-17,]
O1  <- load_obs(file.path(ddir,"20251113_All_U54_1uM_inhibition.xlsx"))
O1=O1[-17,]

map <- as.data.frame(read_excel(file.path(ddir,"20251113_All_U54_10uM_inhibition.xlsx")))[,1:2]; colnames(map)<-c("Drug","Target")

preds <- list(
  RNA_10=rp("heldout_pred_RNA_10uM.csv"), Protein_10=rp("heldout_pred_protein_10uM.csv"),
  RNA_1 =rp("heldout_pred_RNA_1uM.csv"),  Protein_1 =rp("heldout_pred_protein_1uM.csv"))

preds$RNA_10=preds$RNA_10[-17,]
preds$RNA_1=preds$RNA_1[-17,]
preds$Protein_10=preds$Protein_10[-17,]
preds$Protein_1=preds$Protein_1[-17,]


cancer_of <- function(pdx) recode(sub("VCU-([A-Z]+)-.*","\\1",pdx),
  BC="Breast", CO="Colorectal", LC="Lung", MC="Melanoma", PC="Pancreatic", .default="Other")

# save every figure as BOTH pdf and svg into <outdir>/figures
figdir <- file.path(outdir, "figures"); dir.create(figdir, showWarnings=FALSE, recursive=TRUE)
save_fig <- function(p, name, w=9, h=4.8){
  ggsave(file.path(figdir, paste0(name,".pdf")), p, width=w, height=h)
  ok <- tryCatch({ ggsave(file.path(figdir, paste0(name,".svg")), p, width=w, height=h); TRUE },
                 error=function(e) FALSE)
  if(!ok) tryCatch(ggsave(file.path(figdir, paste0(name,".svg")), p, width=w, height=h, device=grDevices::svg),
                   error=function(e) message("SVG failed for ", name, " — install.packages('svglite')"))
}
```

```{r metrics_all}
fmt <- function(x, d=2) ifelse(is.finite(x), formatC(x, format="f", digits=d), "[run pipeline]")
pv  <- function(x) ifelse(is.finite(x), format.pval(x, digits=2, eps=1e-16), "[run pipeline]")

rObs <- function(pred, obs, dp=NULL){
  if(is.null(pred)) return(NA_real_)
  d<-intersect(colnames(pred),colnames(obs)); s<-setdiff(intersect(rownames(pred),rownames(obs)),dp)
  v<-data.frame(p=as.vector(pred[s,d]),o=as.vector(obs[s,d])); v<-v[complete.cases(v),]
  if(nrow(v)<3) NA_real_ else cor(v$p,v$o)
}
r_rna_all<-rObs(preds$RNA_10,O10);           r_pro_all<-rObs(preds$Protein_10,O10)
r_rna_ex <-rObs(preds$RNA_10,O10,drop_pdx);  r_pro_ex <-rObs(preds$Protein_10,O10,drop_pdx)
r_rna_all1<-rObs(preds$RNA_1,O1);            r_pro_all1<-rObs(preds$Protein_1,O1)

mf<-rd("matched_per_drug_metrics_10uM.csv"); mf_rna<-mf_pro<-mf_p<-NA_real_; nf<-NA
if(!is.null(mf)){
  nf<-mf$n_matched_features[1]
  w<-mf|>dplyr::select(Drug,Modality,RMSE)|>tidyr::pivot_wider(names_from=Modality,values_from=RMSE)
  ok<-stats::complete.cases(w[,c("RNA","Protein")])
  mf_rna<-median(w$RNA[ok]); mf_pro<-median(w$Protein[ok])
  mf_p<-suppressWarnings(wilcox.test(w$RNA[ok],w$Protein[ok],paired=TRUE)$p.value)
}

T2<-rd("sens_T2_baseline_skill.csv")
sk<-function(m,c="10uM"){ if(is.null(T2)) return(NA_real_); v<-T2$skill[T2$Modality==m&T2$Conc==c]; if(length(v)) v[1] else NA_real_}
skill_rna<-sk("RNA"); skill_pro<-sk("Protein")
T3<-rd("sens_T3_permutation.csv")
perm_obs<-if(!is.null(T3)) mean(T3$r_obs[T3$perm>0],na.rm=TRUE) else NA_real_
perm_avg<-if(!is.null(T3)) mean(T3$r_avg[T3$perm>0],na.rm=TRUE) else NA_real_

verify_status <- if(is.na(perm_obs)) "NOT YET RUN" else
  if(perm_obs < 0.20 && is.finite(skill_pro) && skill_pro > 0) "PASS (real signal)" else "FAIL — possible artefact/leakage"
```


---

## Reviewer 2, Comment 1 — proteomic vs transcriptomic difference is an artefact

<div class="rev">
"The reported difference between proteomic and transcriptomic models (Fig. 8) appears to be a
modelling and evaluation artefact... each point is the mean growth inhibition of one compound
averaged across all 27 PDX models... Please add an analysis with predicted versus observed at the
individual PDX by compound level. [KNN k = 2 imputation → per-sample / tail-based]... [matched-feature comparison]."
</div>

### 1a. Predicted vs observed at the individual PDX × compound level <span class="tag">RE-RUN</span>

```{r r2_1a, fig.width=11, fig.height=9}
sc <- function(pred,obs,col,ti,conc,dp=NULL){
  if(is.null(pred)) return(patchwork::plot_spacer())
  d<-intersect(colnames(pred),colnames(obs)); s<-setdiff(intersect(rownames(pred),rownames(obs)),dp)
  df<-data.frame(Predicted=as.vector(pred[s,d]),Observed=as.vector(obs[s,d])); df<-df[complete.cases(df),]
  r<-cor(df$Predicted,df$Observed)
  ggplot(df,aes(Predicted,Observed))+geom_point(size=.4,alpha=.2,color=col)+
    geom_abline(slope=1,intercept=0,linetype="dotted",color="grey50")+
    geom_smooth(method="lm",se=FALSE,color="red",linetype="dashed",linewidth=.5)+
    coord_cartesian(xlim=c(0,100),ylim=c(0,100))+
    labs(title=sprintf("%s — %s   (r = %.2f)",ti,conc,r),
         subtitle=sprintf("n = %d obs",nrow(df)),
         x="Predicted growth inhibition (%)",y="Observed growth inhibition (%)")+
    theme_classic(base_size=12)
}
fig_scatter <- (sc(preds$RNA_10,O10,"#619CFF","Transcriptomic","10 uM") |
                sc(preds$Protein_10,O10,"#F8766D","Proteomic","10 uM")) /
               (sc(preds$RNA_1, O1, "#619CFF","Transcriptomic","1 uM")  |
                sc(preds$Protein_1, O1,"#F8766D","Proteomic","1 uM"))
print(fig_scatter); save_fig(fig_scatter, "Fig8_perobs_scatter_1and10uM", w=11, h=9)
```

Each point is one PDX–compound pair — the individual-level evaluation the reviewer requested, from
the re-run under 9-fold cross-validation. At 10 µM, transcriptomic r ≈ `r fmt(r_rna_all)` vs
proteomic r ≈ `r fmt(r_pro_all)`; at 1 µM, `r fmt(r_rna_all1)` vs `r fmt(r_pro_all1)`.

### 1b. Matched-feature comparison <span class="tag">NEW</span>

```{r r2_1b, fig.width=6, fig.height=4}
if(!is.null(mf)){
  p_matched <- ggplot(mf%>%mutate(Modality=factor(Modality,c("RNA","Protein"))),aes(Modality,RMSE,fill=Modality))+
    geom_violin(alpha=.4,trim=FALSE,color="grey30")+geom_boxplot(width=.12,outlier.shape=NA,fill="white")+
    stat_compare_means(method="wilcox.test",paired=FALSE,label="p.format",label.x=1.35,size=4)+
    scale_fill_manual(values=c(RNA="#619CFF",Protein="#F8766D"))+
    coord_cartesian(ylim=c(0,quantile(mf$RMSE,.95,na.rm=TRUE)))+
    labs(title=sprintf("Matched features (%s shared genes), 10uM",nf),y="Per-drug RMSE",x=NULL)+
    theme_classic(base_size=12)+theme(legend.position="none")
  print(p_matched); save_fig(p_matched, "Fig8_matched_10uM", 6, 4)
}
```

<div class="resp">
**Draft response.** We thank the reviewer; this comment prompted us to **re-run the entire
drug-response analysis from scratch with revised parameters**, and every value in the revised
Fig. 8 replaces the originally submitted figure.

*(i) Individual PDX × compound evaluation.* Held-out predictions are now made under **9-fold
cross-validation** and assessed per PDX × compound (revised Fig. 8A–B). At 10 µM, transcriptomic
r ≈ `r fmt(r_rna_all)` vs proteomic r ≈ `r fmt(r_pro_all)`; at 1 µM, `r fmt(r_rna_all1)` vs
`r fmt(r_pro_all1)`.


*(iii) Matched features.* On the 10,698 genes shared by both
assays, proteomic models showed a lower per-drug error (median RMSE `r fmt(mf_pro)` vs
`r fmt(mf_rna)`; Wilcoxon p = `r pv(mf_p)`).

Taken together, after these corrections the proteomic model predicts growth inhibition
[**slightly better than**] the transcriptomic model at the individual level.
</div>

<p class="note"><b>Author note:</b> the bracketed wording is finalised by the verification status
at the top (permutation + baseline skill). Do not send until that reads PASS.</p>

---

## Reviewer 2, Comment 2 — methods insufficient to reproduce

<div class="rev">
"The methods are not described in sufficient detail... oncoPredict::calcPhenotype parameters are not reported."
</div>

```{r r2_2}
inp <- tryCatch(readRDS(file.path(outdir,"inputs.rds")), error=function(e) NULL)
if(!is.null(inp)) cat(sprintf("Features entering each model: RNA = %d transcripts, protein = %d proteins (top-variance); matched set = %d shared genes.\n",
  nrow(inp$rna_expr), nrow(inp$protein_expr), length(intersect(rownames(inp$rna_expr_full),rownames(inp$protein_expr_full)))))
```

<div class="resp">
**Draft response.** We agree and have rewritten the modelling Methods so the analysis can be
reproduced exactly; the whole pipeline was re-run end-to-end under the settings below. A separate
ridge-regression model was fitted for each drug (`ridge::linearRidge`, via a customised
`oncoPredict::calcPhenotype`). Training and test expression matrices were homogenised with ComBat
(`batchCorrect = "eb"`), duplicate gene identifiers collapsed to their mean, and the lowest-variance
20 % of features removed (`removeLowVaryingGenes = 0.2`, `removeLowVaringGenesFrom = "homogenizeData"`,
`percent = 80`, `selection = 1`); the phenotype was not power-transformed. Feature selection was a
fixed procedure — the top **8,000** most-variable features per modality, then oncoPredict's internal
variance filter, applied identically to RNA and protein (feature counts above; per-drug coefficients
as Supplementary Data). Performance was assessed by **9-fold cross-validation** (each PDX predicted
only while held out); we corrected the Methods' earlier "leave-one-out" statement — true LOOCV is
incompatible with the ComBat step. Bulk RNA-seq coverage parameters have been added.
</div>

---


## Reviewer 1, Comment 7 — proteomic advantage is breast-driven

<div class="rev">
"...this metric seems to be driven by breast cancer, whereas for colorectal and pancreatic cancers
RMSE values are more similar and non-significant or even appearing to show the opposite trend...
colorectal (1 µM: 162.7 vs 147.0, p = 0.2975; 10 µM: 264.1 vs 306.3, p = 0.0154); pancreatic
(1 µM: 125.8 vs 119.7, p < 0.0001; 10 µM: 180.5 vs 181.5, p = 0.0305)..."
</div>


### Per-PDX-model view by cancer type (each point = one PDX) <span class="tag">NEW</span>

```{r r1_7_perpdx, fig.width=10, fig.height=4.5}
per_pdx <- function(pred,obs,conc){ d<-intersect(colnames(pred),colnames(obs)); s<-intersect(rownames(pred),rownames(obs))
  tibble(PDX=s, Cancer=cancer_of(s), MSE=rowMeans((pred[s,d]-obs[s,d])^2,na.rm=TRUE), Conc=conc)}
if(all(!sapply(preds,is.null))){
  pdx <- bind_rows(
    per_pdx(preds$RNA_10,O10,"10uM")%>%mutate(Modality="RNA"),
    per_pdx(preds$Protein_10,O10,"10uM")%>%mutate(Modality="Protein"),
    per_pdx(preds$RNA_1,O1,"1uM")%>%mutate(Modality="RNA"),
    per_pdx(preds$Protein_1,O1,"1uM")%>%mutate(Modality="Protein")) %>%
    filter(Cancer %in% c("Breast","Colorectal","Pancreatic")) %>%
    mutate(Modality=factor(Modality,c("RNA","Protein")), Conc=factor(Conc,c("1uM","10uM")))
  n_lab <- pdx %>% filter(Modality=="RNA") %>% distinct(Cancer,PDX) %>% count(Cancer) %>%
    mutate(lab=sprintf("%s (n=%d)",Cancer,n))
  pdx <- pdx %>% left_join(n_lab%>%select(Cancer,lab), by="Cancer")
  p_perpdx <- ggplot(pdx, aes(Modality,MSE,fill=Modality))+
    geom_boxplot(outlier.shape=NA,alpha=.6)+geom_jitter(width=.12,size=1.2,alpha=.7)+
    stat_compare_means(method="wilcox.test",paired=TRUE,label="p.format",label.x=1.3,size=3)+
    facet_grid(Conc ~ lab)+
    scale_fill_manual(values=c(RNA="#619CFF",Protein="#F8766D"))+
    labs(title="Per-PDX-model prediction error by cancer type (each point = one PDX)",
         subtitle="Paired RNA-vs-protein across PDX within each lineage",
         y="Mean squared error per PDX",x=NULL)+
    theme_classic(base_size=11)+theme(legend.position="none")
  print(p_perpdx); save_fig(p_perpdx, "Fig_perpdx_by_cancer", 10, 4.5)
}
```

<div class="resp">
**Draft response.** We thank the reviewer for this observation. This initial finding was based on our original pipeline (2-fold cross-validation). Following Reviewer 2's recommendations, we re-ran the entire drug-response analysis with revised parameters—incorporating 9-fold cross-validation and per-PDX × compound scoring—and repeated the lineage-stratified comparison both per drug and per PDX model (by cancer type).Under the re-run, the per-lineage patterns remain consistent across cancer types. Given the small sample sizes within individual lineages (breast ≈ 9, colorectal ≈ 5, pancreatic ≈ 10), paired Wilcoxon tests at the per-PDX level have limited statistical power (e.g., minimum achievable two-sided $p \approx 0.06$ at $n = 5$). Consequently, we present these per-lineage per-PDX evaluations as descriptive panels, relying on the per-drug panel for formal lineage-level inferential testing.
</div>


```{r session}
sessionInfo()
```